Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-03-23 Ben_Manuscripts/hdphmm/Outline.tex
Updates to methods

  \subsection{The Infinite State Hidden Markov Model}\label{method:IHMM}
  %BJC: There is a fair amount of detail here just so I could understand better by writing it out

  Hidden Markov models (HMMs) are a useful and widely used technique
  for modeling sequences of observations where the probability of the next observation
  in a sequence depends, at least in, part on a previous unobserved, latent or hidden, state.~\cite{beal_infinite_2002}
  %MRS2: it can depend on other things, but ALSO depends on the latent state.
  %BJC2: added 'at least in part
  \begin{itemize}
    \item In the context of our simulations, the observations correspond to 
    the center of mass coordinates of the solutes versus time, and the states
    correspond to the dynamical behavior which give rise to those types
    of observations.
    \item Unfortunately, standard HMMs require the number of hidden states to be known
    a priori.
    \item One can partially overcome this by testing a range of numbers of 
    hidden states and determining which is the best representation of the
    data.
  \end{itemize}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  % and it doesn't do a great job
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, H, a Dirichlet process 
    (DP) generates distributions over a countably infinite number of 
    probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the concentration 
    parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    \item The concentration parameter expresses one's confidence in H relative to the posterior 
    and is closely related to the number of data observations.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

  %BJC: I do need to include this
  \noindent We describe the dynamics of each state using a first order vector autoregressive
   (VAR(1)) model. 
  \begin{itemize}
  	\item In general, a VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  	\label{eqn:var}
  	\end{equation}
  	Previous observations are weighted by coefficient matrices $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
        %MRS2: mean is zero in Z, is it necessarily in R (I realize we've talked about this before, but wanted to get the clarification in text)
    %BJC2: The mean I'm talking about below is different than you are describing above.
    % The multivariate Gausian noise is coming from N(0, \Sigma)
  	\item We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  	covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$.
  	\item We used a conjugate matrix-normal inverse-Wishart prior on parameters
  	$A$ and $\Sigma$ and a conjugate Gaussian prior on $\mathbf{c}$ in order to analytically
  	draw from the posterior.~\cite{fox_nonparametric_2009}
  \end{itemize}

@@

  %BJC2: this is hard to word clearly. Maybe a small figure would help
  We first applied the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$.
    \item By using 2 dimensions, rather than a single radial dimension, we
    increase the likelihood of finding more states since a single radial
    mean can be represented by an infinite number of $x$ and $y$ coordinate combinations.
  \end{itemize}

@@

  The states predicted by the IHMM are heavily influenced by the Gaussian prior
  placed on $c$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The $A$ and $\mathbf{e}_t$ matrices do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $c$ in an intelligent way.
    \item The prior parameters should be chosen such that the means of 
    each state lie within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to be halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item We defined the maximum and minimum to be 2 standard deviations from
    the mean in order to parameterize the variance in each dimension.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen.
  \end{itemize}

  %BJC2: I think it would be useful to continue using this trajectory to illustrate 
  % the parameterization process. It might be bunch of plots stacked on each other.
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{prior_guesses.pdf}
  \caption{The parameters of the prior on $c$ (black line) should be chosen such
  that the means of each state identified in the trajectory (blue line) lie within
  regions of the prior with reasonable probability. We chose the mean of the prior 
  as halfway between the maximum and minimum of each trajectory dimension. We also
  defined the maximum and minimum to be 2 standard deviations from the mean in 
  order to parameterize $\sigma$.}\label{fig:prior_guesses}
  \end{figure}

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequence and parameters for each state.
  \begin{itemize}  
    \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We averaged the parameters from each iteration after the equilibration
    time point.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    \item Since all of the parameters are multidimensional, we found the 
    equilibration time point of each component and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

  % BJC2: It's probably possible to convert to from x, y, z to r, z but
  % I need to figure out the transformation of A. This approach works though.
  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because the mean in $r$ is 
  a useful clustering variable.
  \begin{itemize}
    \item It will also be clearer to analyze the final parameter set in
    terms of $r$ rather than $x$ and $y$.
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item In order to keep the same states found by applying the algorithm
    in 3D, we reapplied the IHMM to the cylindrical trajectories, only allowing
    inference on the VAR(1) parameters with a fixed state sequence. 
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
   \item For each solute studied, we identified between 200-300 independent sets of
   parameters. % BJC2: just a guess for now
   \item Many of these states exhibit very similar dynamical behavior except their
   mean levels are different, especially in the axial direction where solutes could
   get trapped along a broad and continuous range of $z$ coordinates.
   \item Since we do not know the number of states with shared dynamical behavior
   beforehand, we used a non-parametric Bayesian Gaussian mixture model in order
   to group them. % cite scikit-learn  
   \item This is necessary for us to perform any kind of mechanistic speculation.
  \end{itemize}

  We clustered based on 5 variables: the radial means of each state, the two
  eigenvalues of $A$ and the two eigenvalues of $e_t$.
  \begin{itemize}
   \item Unfortunately, the number of independent parameter sets is too low for
   clustering in higher dimensions. 
   \item Ideally, we could use all of the entries of $A$ and $e_t$ in addition to 
   the radial means.
   \item It would not be helpful to use the means in the $z$ direction because solutes
   are essentially unbound in that direction, while there tends to be radially
   dependent behavior as shown in our previous work.~\cite{coscia_chemically_2019}  % cite new paper too.
  \end{itemize}

  We remapped the state sequence based on the clustered assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states.
  \end{itemize}

  We obtained clustered $\mathbf{c}$ vectors based on the means of 
  $\mathbf{c}$ belonging to the same cluster.
  \begin{itemize}
   \item We only care about the $r$ component of $c$ because solute
   trajectories are not boudn in the $z$ direction.
  \end{itemize}

  We used the IHMM algorithm in order to infer $A$ and $e_t$ of the 
  clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $e_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on modified this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations by drawing state sequences 
  based on the rows of $T$.
  \begin{itemize}
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item We set the unconditional mean of each state based on the position 
    before the state transition occurred.
  \end{itemize}

>>>>

  \subsection{The Infinite State Hidden Markov Model}\label{method:IHMM}

  Hidden Markov models (HMMs) are a useful and widely used technique
  for modeling sequences of observations where the probability of the next observation
  in a sequence depends, at least in part, on a previous unobserved, latent or hidden, state.~\cite{beal_infinite_2002}
  %MRS2: it can depend on other things, but ALSO depends on the latent state.
  %BJC2: added 'at least in part
  \begin{itemize}
    \item In the context of our simulations, the observations correspond to 
    the center of mass coordinates of the solutes versus time, and the states
    correspond to the dynamical behavior which give rise to those types
    of observations.
    \item The probability of transitioning to a state based on the current
    state is mathematically defined in terms of an $n\times n$ transition
    probability matrix, $T$, where $n$ is the number of states.
    \item Unfortunately, standard HMMs require $n$ to be known
    \textit{a priori}.
    \item One can partially overcome this by testing a range of numbers of 
    hidden states and determining which is the best representation of the
    data.
  \end{itemize}

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, H, a Dirichlet process 
    (DP) generates distributions over a countably infinite number of 
    probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the concentration 
    parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    \item The concentration parameter expresses one's confidence in H relative to the posterior 
    and is closely related to the number of data observations.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

  \noindent We describe the dynamics of each state using a first order vector autoregressive
   (VAR(1)) model. 
  \begin{itemize}
  	\item In general, a VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  	\label{eqn:var}
  	\end{equation}
  	Previous observations are weighted by coefficient matrices $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
        %MRS2: mean is zero in Z, is it necessarily in R (I realize we've talked about this before, but wanted to get the clarification in text)
    %BJC2: The mean I'm talking about below is different than you are describing above.
    % The multivariate Gausian noise is coming from N(0, \Sigma)
  	\item We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  	covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$.
  	\item We used a matrix-normal inverse-Wishart prior on parameters $A$ and $\Sigma$ 
  	and a Gaussian prior on $\mathbf{c}$ in order to infer their values.~\cite{fox_nonparametric_2009}
  	%BJC2: I'm not sure above is enough to make it clear how I got the parameters. Maybe I should
  	% explicitly refer to the fox paper above which outlines the inference procedure in more detail?
  \end{itemize}

  % BJC2: It might be useful to walk through the following procedure with an example 
  % trajectory

@@

  We first applied the IHMM algorithm to 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    \item By using 2 dimensions, rather than a single radial dimension, we
    increase the likelihood of finding more states since a single radial
    mean can be represented by an infinite number of $x$ and $y$ coordinate combinations.
  \end{itemize}

@@

  The states predicted by the IHMM are heavily influenced by the Gaussian prior
  placed on $c$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of the $A$ and $\mathbf{e}_t$ matrices do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the means of 
    each state lie within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to be halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the maximum and minimum
    to be 2 standard deviations from the mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

  %BJC2: I think it would be useful to continue using this trajectory to illustrate 
  % the parameterization process. It might be bunch of plots stacked on each other.
  %BJC2: More labels might be useful if not cluttered
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{prior_guesses.pdf}
  \caption{The parameters of the prior on $c$ (black line) should be chosen such
  that the mean levels of each state identified in the trajectory (blue line) lie within
  regions of the prior with reasonable probability. We chose the mean of the prior 
  as halfway between the maximum and minimum of each trajectory dimension. We chose 
  $\sigma$ of the prior by defining the maximum and minimum to be 2 standard deviations
  from the mean (dashed lines).}\label{fig:prior_guesses}
  \end{figure}

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We averaged the parameters from each iteration recorded after the
    equilibration time point.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    \item Since all of the parameters are multidimensional, we found the 
    equilibration time point of each component and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

  % BJC2: It's probably possible to convert to from x, y, z to r, z but
  % I need to figure out the transformation of A. This approach works though.
  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because the mean in $r$ is 
  a useful clustering variable.
  \begin{itemize}
    \item It will also be clearer to analyze the final parameter set in
    terms of $r$ rather than $x$ and $y$.
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item In order to keep the same states found when we applied the algorithm
    in 3D, we reapplied the IHMM to the cylindrical trajectories with a fixed 
    state sequence, only allowing inference on the VAR(1) parameters. 
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
   \item For each solute studied, we identified between 200-300 independent sets of
   parameters. % BJC2: just a guess for now
   \item Many of these states exhibit very similar dynamical behavior except their
   mean levels are different, especially in the axial direction where solutes could
   get trapped along a broad and continuous range of $z$ coordinates.
   \item Since we do not know the number of states with shared dynamical behavior
   beforehand, we used a non-parametric Bayesian Gaussian mixture model in order
   to group them. % cite scikit-learn  
%   \item This is necessary for us to perform any kind of mechanistic speculation.
  \end{itemize}

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on 5 variables: the radial means of each state, the two
  eigenvalues of $A$ and the two eigenvalues of $e_t$.
  \begin{itemize}
   \item Unfortunately, the number of independent parameter sets is too low for
   clustering in higher dimensions. 
   \item Ideally, we could use all of the entries of $A$ and $e_t$ in addition to 
   the radial means.
   \item It would not be helpful to use the means in the $z$ direction because solutes
   are essentially unbound in that direction, while 
   \item However, in our previous work, we observed radially dependent dynamical 
   behavior.~\cite{coscia_chemically_2019}  % cite new paper too.
  \end{itemize}

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states.
  \end{itemize}

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of $\mathbf{c}$ assigned to the same cluster.
  \begin{itemize}
   \item We only care about the $r$ component of $\mathbf{c}$ because solute
   trajectories are not bound in the $z$ direction.
  \end{itemize}

  We used the IHMM algorithm in order to infer $A$ and $\mathbf{e}_t$ of the 
  clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    \item We drew state sequences with transition probabilities weighted by the 
    the rows of $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}
